A Switching Planner for Combined Task and Observation Planning

### 2010-11-20 code/subarchitectures/planner.sa/trunk/src/latex/ICAPS-2011/moritz_2011.tex
Better POMDP section.
 - Charles

We model a process state $\state$ as the set of propositions that are
true of the state. Notationally, we have $\state \subseteq
\props$. State change is induced by application of actions. A
stochastic action $\stochAction \in \stochActions$ is applicable if
its precondition $\poss(\stochAction)$, a set of propositions, are
satisfied in the current state -- I.e., $\poss(\stochAction) \subseteq
\state$. We write $\stochActions(\state)$ for set of actions
applicable at state $\state$.  If action $\stochAction \in
\stochActions(\state)$ is applied at \state, nature chooses one
element amongst a small set of deterministic outcomes
$\detActions(\stochAction) \equiv \{\detAction_1, \ldots, \detAction_k
\}$. We denote $\mu_{\stochAction}(\detAction_i)$ the probability that
nature takes outcome $\detAct_i$, and for all \stochAction\ we require
$\sum_{\detAction_i \in \detActions(\stochAction)}
\mu_{\stochAction}(\detAction_i) = 1$. The chosen outcome has an
effect on the state given in terms of two lists of propositions. The
add-list $\add(\detAction)$ and delete-list
$\delete(\detAction)$.\footnote{If a proposition is in the add-list of
$\detAction$, then it cannot be in the delete-list and vice versa.}
If outcome $\detAction$ with $\add(\detAction) := [\prop_1,
..,\prop_n]$ and $\delete(\detAction) := [\prop_1, ..,\prop_m]$ is
chosen by nature when \stochAction\ is applied at state $\state$, then
the resultant state is $ ( \state \cup \add(\detAction) ) \backslash
\delete(\detAction)$ -- i.e., propositions from $\add(\detAction)$ are
added to $\state$, and those from $\delete(\detAction)$ are removed
from $\state$.

We are concerned with problems that feature partial
observability. These have a perceptual model given in terms of a
finite set of stochastic {\em senses} $\stochSenses$, deterministic
sensing outcomes $\detSenses$, and perceptual propositions
$\percepts$, called {\em percepts}. Here, an observation $\observ$ is
a set of percepts $\observ \subseteq \percepts$, and we denote
\observations\ the set of reachable observations. The underlying state
of the process cannot be observed directly, rather, senses
$\stochSense \in \stochSenses$ effect an observation $\observ \in
\observations$ that informs what should be believed about the state a
process is in. In detail, if $\action$ is applied effecting a
transition to a successor state $\state'$, then an observation occurs
according to the active senses $\stochSenses(\stochAction, \state')
\subseteq \stochSenses$. A percept is active, written $\stochSense \in
\stochSenses(\stochAction, \state')$, if the senses'
action-precondition, $\poss_\stochActions(\stochSense)$, is equal to
$\stochAction$, and the state-precondition $\poss_\states(\stochSense)
\subseteq \propositions$ is satisfied by the state $\state'$; In the
usual sense that $\poss_\states(\stochSense) \subseteq \state'$.
%%
When a sense is active, nature must choose exactly one outcome amongst
a small set of deterministic choices $\detSenses(\stochSense)
\equiv \{\detSense_1, \ldots, \detSense_k \}$, so that for each
$i$ we have $\detSense_i \subseteq \percepts$. The probability of
the $i^{th}$ element being chosen is given by
$\psi_{\stochSense}(\detSense_i)$, where $\sum_{\detSense_i \in
\detSenses(\stochSense)} \psi_{\stochSense}(\detSense_i) =
1$. The observation received by the agent corresponds to the union of
perceptual propositions from chosen elements of active senses.

@@

Finally, we make a number of fairly standard assumptions. First, that
action execution and sensing occurs instantaneously, and that only one
action can be applied at a plan-step. Second, it can arise that in
some propositional states an action is applicable, and in others it is
not. Moreover, a belief-state can assign non-zero probability to
states where that applicability holds, and states where it does
not. We differ
from~\citeauthor{younes:littman:04}~(\citeyear{younes:littman:04}),
because we forbid the execution of actions that are not applicable in
any state of the current belief.  Also,
unlike~\citeauthor{hoffmann:brafman:2006}~(\citeyear{hoffmann:brafman:2006}),
we do incorporate a PPDDL-like default semantics, supposing that when
an action \stochAction\ is executed at $\state_i$ where
$\bstate(\state_i) > 0$, that it has no effect on states $\state_i$ if
$\poss(\stochAction) \not\subseteq \state_i$ and $\bstate(\state_i) >
0$.

\subsection{Costs, Rewards, and Plan Evaluation}

Whereas until now we have considered a POMDP model that is factored in
terms of propositions and percepts, in order to discuss utilities and
policies it is convenient to consider the underlying decision process
in a flat format. This underlying decision process is given by the
tuple $\langle \states, \bstate_0, \actions, \transProb, \reward,
\observations, \obsDist \rangle$. Here $\bstate_0$ is the initial
belief-state, \states\ is the finite set of reachable propositional
states, \actions\ is the finite set of actions, and \observations\ is
the finite set of reachable observations (i.e., perceptual states).
Where $\state, \state' \in \states$, $a \in \actions$, from $\mu$ we
have a state transition function $\transProb(\state, \action,
\state')$ giving the probability of a transition from state $s$ to
$s'$ if $a$ is applied. For any $\state$ and $\action$ we have
$\sum_{\state' \in \states} \transProb(\state, \action, \state') = 1$.
%%
Function $\reward:\states \times \actions \to \Re$ is a bounded real
valued reward function.  Consequently there is a positive constant $c$
so that for all $\state \in \states$ and $\action \in \actions$,
$|\reward(\state, \action)| < c$.
%%
From $\psi$ we have that for each $\state \in \states$ and action
$\action \in \actions$, an observation $\observ \in \observations$ is
generated independently according to a probability distribution
$\obsDist(\state, \action)$. We denote $\obsDist_\observ(\state,
\action)$ the probability of getting observation $\observ$ in state
$\state$. For $\state$ and $\action$ we have $\sum_{\observ \in
\observations} \obsDist_\observ(\state, \action) = 1$.

Successive state estimation is available if we applying Bayes' rule.
Taking the current belief $\bstate$ as our {\em prior}, and supposing
action $\action$ is executed with perceptive outcome $\observ$, then
the probability that we are in $\state$ in the successive belief-state
$\bstate'$ is given by:

@@

A general formalism for representing solutions to POMDPs is the
finite-state controller (FSC). This is a three-tuple $\langle \nodes,
\selection, \transition \rangle$ where: $\node \in \nodes$ is a set of
nodes, $\selection_\node(\action) = P(\action | \node)$, and
$\transition_\node(\action, \observation, \node') = P(\node'|\node,
\action, \observation)$. The value of state $\state$ at node $\node$
given a problem model, written $V_\node(\state)$, is:

\begin{equation}\label{eq:evaluation}
\begin{array}{lcl}
V_\node(\state) & = & \sum_{\action \in \actions}
\selection_\node(\action) \reward(\state, \action) \;\; + \vspace{1ex} \\

&& \hspace{-10ex} \beta \sum_{\action, \observation,
\state', \node'} \transition_\node(\action, \observation, \node')
\transProb(\state, \action, \state') \obsDist_\observ(\state',
\action) V_{\node'}(\state')
\end{array}
\end{equation}

\noindent The value of $\bstate$ according to a controller is
therefore:

\begin{equation} \label{eq:valueBelief}
V_{\pp{FSC}}(\bstate) = \max_{\node \in \nodes} \sum_{\state \in \states} \bstate(\state) V_\node(\state)
\end{equation}

@@

\small{
\begin{tabtt}
(\=:functions  ;; state fluents\\
  \> (is-in ?v - visual-object) - location )\\
(:observable-functions  ;; perceptual fluents\\
  \> (o-is-in ?v - visual-object) - location )
\end{tabtt}
}

@@

\small{
\begin{tabtt}
(\=:action look-for-object \+ \\
   :parameters (\=?r - robot ?v - visual-object\\
   \> ?l - location) \\
   :precondition (and (= (is-in ?r) ?l) ) \\
   :effect (and (assign (reward) -3) ) ) \\
\end{tabtt}
}

@@

\small{
\begin{tabtt}
(\= :sense vision \+\\
 :parameters \= (\= ?r - robot ?v - visual-object\\
 \>\>  ?l - location) \\
 :execution \> ( \> look-for-object ?r ?v ?l) \\
 :precondition (and (= (is-in ?r) ?l) ) \\
 :effect \>  (  \> and (when (= (is-in ?v) ?l) \\
   \> \> (probabilistic 0.8 \\
   \>  \>(= (o-is-in ?v) ?l))) \\
  \> (when (not (= (is-in ?v) ?l)) \\
   \>  \> (probabilistic 0.1 \\
   \>  \> (= (o-is-in ?v) ?l))))) \\
\end{tabtt}
}

@@

\small{
\begin{tabtt}
(\=:init (= (is-in R2D2) kitchen) \+ \\
       (probabilistic \=.8 (= (is-in box) office)  \\
		      \>.2 (= (is-in box) kitchen)) \\
       (probabilistic .3 (= (is-in cup) office)  \\
		      \>.7 (= (is-in cup) kitchen))) \\
\end{tabtt}
}

@@

\begin{tabular}{cccc}
\hline
Probability & (is-in R2D2)  & (is-in box)  & (is-in cup) \\
\hline
.24 & kitchen & office & office \\
.06 & kitchen & kitchen & office \\
.56 & kitchen & office & kitchen \\
.14 & kitchen & kitchen & kitchen \\
\hline
\end{tabular}

>>>>

We model a process state $\state$ as the set of propositions that are
true of the state. Notationally, we have $\state \subseteq
\props$. State change is induced by application of actions. A
stochastic action $\stochAction \in \stochActions$ is applicable if
its precondition $\poss(\stochAction)$, a set of propositions, are
satisfied in the current state -- I.e., $\poss(\stochAction) \subseteq
\state$. We write $\stochActions(\state)$ for set of actions
applicable at state $\state$.  If action $\stochAction \in
\stochActions(\state)$ is applied at \state, nature chooses one
element amongst a small set of deterministic outcomes
$\detActions(\stochAction) \equiv \{\detAction_1, \ldots, \detAction_k
\}$. We denote $\mu_{\stochAction}(\detAction_i)$ the probability that
nature takes outcome $\detAct_i$, and for all \stochAction\ we require
$\sum_{\detAction_i \in \detActions(\stochAction)}
\mu_{\stochAction}(\detAction_i) = 1$. The chosen outcome has an
effect on the state given in terms of two lists of propositions. The
add-list $\add(\detAction)$ and delete-list
$\delete(\detAction)$.\footnote{If a proposition is in the add-list of
$\detAction$, then it cannot be in the delete-list and vice versa.}
If outcome $\detAction$ with $\add(\detAction) := [\prop_1,
..,\prop_n]$ and $\delete(\detAction) := [\overline\prop_{n+1},
..,\overline\prop_m]$ is chosen by nature when \stochAction\ is
applied at state $\state$, then the resultant state is $ ( \state \cup
\add(\detAction) ) \backslash \delete(\detAction)$ -- i.e.,
propositions from $\add(\detAction)$ are added to $\state$, and those
from $\delete(\detAction)$ are removed from $\state$.

We are concerned with problems that feature partial
observability. These have a perceptual model given in terms of a
finite set of stochastic {\em senses} $\stochSenses$, deterministic
sensing outcomes $\detSenses$, and perceptual propositions
$\percepts$, called {\em percepts}. Here, an observation $\observ$ is
a set of percepts $\observ \subseteq \percepts$, and we denote
\observations\ the set of reachable observations. The underlying state
of the process cannot be observed directly, rather, senses
$\stochSense \in \stochSenses$ effect an observation $\observ \in
\observations$ that informs what should be believed about the state a
process is in. In detail, if $\action$ is applied effecting a
transition to a successor state $\state'$, then an observation occurs
according to the active senses $\stochSenses(\stochAction, \state')
\subseteq \stochSenses$. A sense $\stochSense$ is active, written
$\stochSense \in \stochSenses(\stochAction, \state')$, if the senses'
action-precondition, $\poss_\stochActions(\stochSense)$, is equal to
$\stochAction$, and the state-precondition $\poss_\states(\stochSense)
\subseteq \propositions$ is satisfied by the state $\state'$; In the
usual sense that $\poss_\states(\stochSense) \subseteq \state'$.
%%
When a sense is active, nature must choose exactly one outcome amongst
a small set of deterministic choices $\detSenses(\stochSense)
\equiv \{\detSense_1, \ldots, \detSense_k \}$, so that for each
$i$ we have $\detSense_i \subseteq \percepts$. The probability of
the $i^{th}$ element being chosen is given by
$\psi_{\stochSense}(\detSense_i)$, where $\sum_{\detSense_i \in
\detSenses(\stochSense)} \psi_{\stochSense}(\detSense_i) =
1$. The observation received by the agent corresponds to the union of
perceptual propositions from chosen elements of active senses.

@@

Finally, we make a number of fairly standard assumptions. First, that
action execution and sensing occurs instantaneously, and that only one
action can be applied at a plan-step. Second, it can arise that in
some propositional states an action is applicable, and in others it is
not. Moreover, a belief-state can assign non-zero probability to
states where that applicability holds, and states where it does
not. We differ
from~\citeauthor{younes:littman:04}~(\citeyear{younes:littman:04}),
because we forbid the execution of actions that are not applicable in
any state of the current belief.  Also,
unlike~\citeauthor{hoffmann:brafman:2006}~(\citeyear{hoffmann:brafman:2006}),
we do incorporate a PPDDL-like default semantics, supposing that when
an action \stochAction\ is executed at $\state_i$, that it has no
effect on states $\state_i$ if $\poss(\stochAction) \not\subseteq
\state_i$.

\subsection{Costs, Rewards, and Belief Revision}

Whereas until now we have considered a POMDP model that is factored in
terms of propositions and percepts, in order to discuss utilities and
policies it is convenient to consider the underlying decision process
in a flat format. This underlying decision process is given by the
tuple $\langle \states, \bstate_0, \actions, \transProb, \reward,
\observations, \obsDist \rangle$. Here $\bstate_0$ is the initial
belief-state, \states\ is the finite set of reachable propositional
states, \actions\ is the finite set of actions, and \observations\ is
the finite set of reachable observations (i.e., perceptual states).
Where $\state, \state' \in \states$, $a \in \actions$, from $\mu$ we
have a state transition function $\transProb(\state, \action,
\state')$ giving the probability of a transition from state $s$ to
$s'$ if $a$ is applied. For any $\state$ and $\action$ we have
$\sum_{\state' \in \states} \transProb(\state, \action, \state') = 1$.
%%
Function $\reward:\states \times \actions \to \Re$ is a bounded real
valued reward function. Therefore a finite positive constant $c$
exists so that for all $\state \in \states$ and $\action \in
\actions$, $|\reward(\state, \action)| < c$. We model costs are
negative rewards.
%%
From $\psi$ we have that for each $\state \in \states$ and action
$\action \in \actions$, an observation $\observ \in \observations$ is
generated independently according to a probability distribution
$\obsDist(\state, \action)$. We denote $\obsDist_\observ(\state,
\action)$ the probability of getting observation $\observ$ in state
$\state$. For $\state$ and $\action$ we have $\sum_{\observ \in
\observations} \obsDist_\observ(\state, \action) = 1$.

In the POMDP model successive state estimation is available by
application of Bayes' rule.  Taking the current belief $\bstate$ as
the {\em prior}, and supposing action $\action$ is executed with
perceptive outcome $\observ$, then the probability that we are in
$\state$ in the successive belief-state $\bstate'$ is given by:

@@

\subsection{Plan Evaluation}

An optimal solution to a finite-horizon POMDP is a contingent plan,
and can be expressed as a mapping from observation histories to
actions. Although suboptimal in general, useful plans can also take a
classical sequential format. This is the case in {\em conformant}
planning, where the objective is to find a sequence of actions that
achieves a goal ---I.e., reaches a state that satisfies some given
Boolean condition--- with probability $1$.  More generally, a useful
formalism for representing solutions to POMDPs is the finite-state
controller (FSC). This is a three-tuple $\langle \nodes, \selection,
\transition \rangle$ where: $\node \in \nodes$ is a set of controller
states, $\selection_\node(\action) = P(\action | \node)$ gives the
probability that the controller prescribes \action\ when in state
\node, and $\transition_\node(\action, \observation, \node') =
P(\node'|\node, \action, \observation)$ gives the probability the
controller transitions to state $\node'$, supposing we execute
\action\ in state \node\ receiving observation \observation. When
acting for finite $N$ steps, the value of a controller corresponds to
the expected cumulative reward:

\[
V_{\pp{FSC}}(\bstate) = \Expect \bigg{[} 
\sum_{t=0}^{N}  \reward(\bstate_t) \mid \pp{FSC}, \bstate = \bstate_0 \bigg{]}
\]

\noindent Where $b_t$ is the belief state at time $t$. Finally, it is
useful to note that there is a corresponding deterministic FSCs for
both sequential (i.e., conformant) and contingent plans.

%% More generally, a useful
%% formalism for representing solutions to POMDPs is the finite-state
%% controller (FSC). This is a three-tuple $\langle \nodes, \selection,
%% \transition \rangle$ where: $\node \in \nodes$ is a set of nodes,
%% $\selection_\node(\action) = P(\action | \node)$, and
%% $\transition_\node(\action, \observation, \node') = P(\node'|\node,
%% \action, \observation)$. Where the value of acting is the discounted
%% accumulated reward over an infinite horizon, then each controller can
%% be assigned a value, written $V_\node(\state)$, expressed in the usual
%% way according to the Bellman equation:

%% \begin{equation}\label{eq:evaluation}
%% \begin{array}{lcl}
%% V_\node(\state) & = & \sum_{\action \in \actions}
%% \selection_\node(\action) \reward(\state, \action) \;\; + \vspace{1ex} \\

%% && \hspace{-10ex} \beta \sum_{\action, \observation,
%% \state', \node'} \transition_\node(\action, \observation, \node')
%% \transProb(\state, \action, \state') \obsDist_\observ(\state',
%% \action) V_{\node'}(\state')
%% \end{array}
%% \end{equation}

%% \noindent Here, the value of {\em prior} $\bstate$ given an FSC is
%% then:

%% \begin{equation} \label{eq:valueBelief}
%% V_{\pp{FSC}}(\bstate) = \max_{\node \in \nodes} \sum_{\state \in \states} \bstate(\state) V_\node(\state)
%% \end{equation}

%% \noindent There is a corresponding deterministic FSCs for both
%% sequential (i.e., conformant) and contingent
%% plans. Eq~\ref{eq:evaluation} gives their value in the finite-horizon
%% case, if we take $\beta = 1$, and alter the process dynamics so there
%% is a compulsory transition to a zero utility sink state when that
%% horizon is reached.

@@

\small
\begin{tabtt}
(\=:functions  ;; state fluents\\
  \> (is-in ?v - visual-object) - location )\\
(:observable-functions  ;; perceptual fluents\\
  \> (o-is-in ?v - visual-object) - location )
\end{tabtt}
\normalsize

@@

\small
\begin{tabtt}
(\=:action look-for-object \+ \\
   :parameters (\=?r - robot ?v - visual-object\\
   \> ?l - location) \\
   :precondition (and (= (is-in ?r) ?l) ) \\
   :effect (and (assign (reward) -3) ) ) \\
\end{tabtt}
\normalsize

@@

\small
\begin{tabtt}
(\= :sense vision \+\\
 :parameters \= (\= ?r - robot ?v - visual-object\\
 \>\>  ?l - location) \\
 :execution \> ( \> look-for-object ?r ?v ?l) \\
 :precondition (and (= (is-in ?r) ?l) ) \\
 :effect \>  (  \> and (when (= (is-in ?v) ?l) \\
   \> \> (probabilistic 0.8 \\
   \>  \>(= (o-is-in ?v) ?l))) \\
  \> (when (not (= (is-in ?v) ?l)) \\
   \>  \> (probabilistic 0.1 \\
   \>  \> (= (o-is-in ?v) ?l))))) \\
\end{tabtt}
\normalsize

@@

\small
\begin{tabtt}
(\=:init (= (is-in R2D2) kitchen) \+ \\
       (probabilistic \=.8 (= (is-in box) office)  \\
		      \>.2 (= (is-in box) kitchen)) \\
       (probabilistic .3 (= (is-in cup) office)  \\
		      \>.7 (= (is-in cup) kitchen))) \\
\end{tabtt}
\normalsize

@@

\small
\begin{tabular}{cccc}
\hline
Probability & (is-in R2D2)  & (is-in box)  & (is-in cup) \\
\hline
.24 & kitchen & office & office \\
.06 & kitchen & kitchen & office \\
.56 & kitchen & office & kitchen \\
.14 & kitchen & kitchen & kitchen \\
\hline
\end{tabular}
\normalsize
